---
title: "RNA-seq-DEseq"
output: html_document
date: '2022-04-15'
---

## RNA-Seq with DEseq2

```{r}
# Install and load packages

if (!requireNamespace("BiocManager", quietly = TRUE)) install.packages("BiocManager")
BiocManager::install("DESeq2")
BiocManager::install("edgeR")
BiocManager::install("biomaRt")
BiocManager::install('PCAtools')
BiocManager::install('EnhancedVolcano')
```

```{r}
###################### load the raw count matrix #######################

setwd("./") #Path_to_working_directory

```

```{r}
######################  Filter for coding genes (In case want to filter non-coding Genes) ########################
library("biomaRt")

mart <- useMart(biomart="ensembl", dataset="hsapiens_gene_ensembl")
all_coding_genes <- getBM(attributes = c( "hgnc_symbol"), filters = c("biotype"), values = list(biotype="protein_coding"), mart = mart)
rawcount<-read.table ("RawGeneCounts.tsv",header=TRUE,  sep="\t",  row.names=1)
rawcount <- rawcount[row.names(rawcount) %in%  all_coding_genes$hgnc_symbol,]
```

```{r}
###################### Data annotation  #################################

anno <-read.table ("Annotation_of_samples.csv",header=TRUE,  sep=",") ##In this case Two coulmns (a) sample (b) Condition
rownames(anno) <- anno$sample
```

# Define conditions (for contrast) that you want to compare if you have more than one #control #case
# This is pair-wise comparison, so only consider one pair at one time

```{r}

firstC<-"case1"       #case1 #case2 #case3 etc          
SecondC <-"Control"     
p.threshold <- 0.05   ##define threshold for filtering

### subset raw and conditional data for defined pairs

anno <- anno[(anno$Condition ==firstC |anno$Condition ==SecondC),]

anno <- anno[anno$sample %in% names(rawcount),]
rawcount <- rawcount[,names(rawcount) %in% anno$sample]
```


```{r}
############################### Create DESeq2 datasets #############################
library(DESeq2)

dds <- DESeqDataSetFromMatrix(countData = rawcount, colData = anno, design = ~Condition )

##dds <- DESeqDataSetFromMatrix(countData = rawcount, colData = anno, design =  ~Batch+Condition )  ###USE this one if you have extra col in anno data with Batch info
#View(counts(dds))

dds <- estimateSizeFactors(dds)

normalized_counts <- counts(dds, normalized=TRUE)  ## However,Deseq2 use raw count #This one is good for visualization purpose

#View(normalized_counts)
write.table(normalized_counts, file="normalized_counts.txt", sep="\t", quote=F, col.names=NA)

### Transform counts for data visualization
rld <- rlog(dds, blind=TRUE)
### Plot PCA 
plotPCA(rld, intgroup="Condition")

## Run DESEQ2
dds <- DESeq(dds)

##ensure your data is a good fit for the DESeq2 model
plotDispEsts(dds)
```

```{r}
################# contrast based  comparison ##########################

#In case of multiple comparisons ## we need to change the contrast for every comparision
contrast<- c("Condition",firstC,SecondC)

res <- results(dds, contrast=contrast)
print(res)
### Valcono plot
library(EnhancedVolcano)

EnhancedVolcano(res,
                lab = rownames(res),
                x = 'log2FoldChange',
                y = 'pvalue')   ## Default cut-off for log2FC is >|2| and for P value is 10e-6. USE  pCutoff = 10e-6, FCcutoff = 2.0 


res$threshold <- as.logical(res$padj < p.threshold)  #Threshold defined earlier

nam <- paste('down_in',firstC, sep = '_')
#res$nam <- as.logical(res$log2FoldChange < 0)
res[, nam] <- as.logical(res$log2FoldChange < 0)

genes.deseq <- row.names(res)[which(res$threshold)]
genes_deseq2_sig <- res[which(res$threshold),]

```

```{r}
########### Plots normalized count of top 20 genes ## sorted based on padjust and filter by |logFC| >=1

res$gene <- row.names(res)
print(res)

# Order results by padj values

#library(dplyr)
library(tidyverse)

top20 <- res %>% 
  as.data.frame %>%
  arrange(padj) %>% 	#Arrange rows by padj values
  filter(abs(log2FoldChange) >=1) %>%   #filter based on logFC
  pull(gene) %>% 		#Extract character vector of ordered genes
  head(n=20) 		#Extract the first 20 genes

top20_norm <- as.data.frame(normalized_counts[rownames(normalized_counts) %in% top20,])

top20_norm_v2 <- top20_norm ## will use later for heatmap

top20_norm <- (top20_norm+1) ## in later step to remove infinity bias due to log

top20_norm$gene <-  row.names(top20_norm)  
top20_norm <- top20_norm %>% 
  pivot_longer(!gene, names_to = "samplename", values_to = "normalized_counts") # Gathering the columns to have normalized counts to a single column

# Create tibbles including row names
mov10_meta <- anno %>% 
  rownames_to_column(var="samplename") %>% 
  as_tibble()

top20_norm <- inner_join(mov10_meta, top20_norm)

################3
## plot using ggplot2

ggplot(top20_norm) +
  geom_point(aes(x = gene, y = normalized_counts, color = Condition)) +
  scale_y_log10() +
  xlab("Genes") +
  ylab("log 10 CPM Normalized Counts") +
  ggtitle("Top 20 Significant DE Genes with abs(logFC) =>1") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(plot.title = element_text(hjust = 0.5))

##################
library(RColorBrewer)
### Set a color palette
heat_colors <- brewer.pal(6, "YlOrRd")

### Run pheatmap
library(pheatmap) 
pheatmap(top20_norm_v2 , 
         color = heat_colors, 
         cluster_rows = T, 
         show_rownames = T,
         annotation_col = anno[,1:2], 
         border_color = NA, 
         fontsize = 10, 
         scale = "row", 
         fontsize_row = 10, 
         height = 20)

```

```{r}
### Add EntrezID column to results dataframe for easier downstream processing ####

genes.deseq.entrezid <- getBM(attributes = c("hgnc_symbol", "entrezgene_id"), filters = "hgnc_symbol", values = row.names(res), mart = mart)
entrez_df = as.data.frame(genes.deseq.entrezid)

res_df = as.data.frame(res)
res_df$hgnc_symbol = row.names(res_df)
row.names(res_df) <- NULL

merged = merge(res_df, entrez_df)
print(merged)

genes_deseq2_sig_df = as.data.frame(genes_deseq2_sig)
genes_deseq2_sig_df$hgnc_symbol = row.names(genes_deseq2_sig_df)
row.names(genes_deseq2_sig_df) <- NULL

sig_merged = merge(genes_deseq2_sig_df, entrez_df)

file <- paste('Deseq2_',firstC,'_v_',SecondC,'_results_significant_padj',p.threshold,'.csv',sep = '')
all_results <- paste('Deseq2_',firstC,'_v_',SecondC,'_all_results.csv',sep = '')

write.table(sig_merged,file,sep = ",")
write.table(merged,all_results,sep = ",")

```

```{r}
######### Rank all genes based on their fold change #########

BiocManager::install("clusterProfiler", force = TRUE)
BiocManager::install("pathview", force = TRUE)
BiocManager::install("enrichplot", force = TRUE)

library(clusterProfiler)
library(enrichplot)
library(ggplot2)

# SET THE DESIRED ORGANISM HERE
organism = "org.Hs.eg.db"

BiocManager::install(organism, character.only = TRUE, force = TRUE)
library(organism, character.only = TRUE)
keytypes(org.Hs.eg.db)
#We will take the log2FoldChange value from previously saved significant results file
#Deseq2_case1_v_Control_results_significant.csv

# reading in data from deseq2
df <- read.csv("Deseq2_case1_v_Control_results_significant_padj0.05.csv")
df
 
#https://learn.gencore.bio.nyu.edu/rna-seq-analysis/gene-set-enrichment-analysis/

# we want the log2 fold change 
original_gene_list <- df$log2FoldChange
print(original_gene_list)

# name the vector
names(original_gene_list) <- df$entrezgene_id

# omit any NA values 
gene_list<-na.omit(original_gene_list)

# sort the list in decreasing order (required for clusterProfiler)
gene_list = sort(gene_list, decreasing = TRUE)

print(gene_list)

```

```{r}
#### Gene Set Enrichment #####

library(stats)

gse <- gseGO(geneList=gene_list, 
             ont ="ALL", 
             keyType = "ENTREZID", 
             minGSSize = 3, 
             maxGSSize = 800, 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = org.Hs.eg.db, 
             pAdjustMethod = "none")

```

We can see that in our dataset not a single value is enriched at a pvalue cut-off of 0.05.

Lets exlore other functions with a sample dataset and see what analysis we can do with a list of differrentially expressed genes from geneList dataset of DOSE package

##GO Enrichment Analysis of a gene set. 
Given a vector of genes, enrichGO function will return the enrichment GO categories after FDR control.

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)
library(enrichplot)
library(GOSemSim)
library(ggnewscale)
library(DOSE)

data(geneList)
View(geneList)
gene <- names(geneList)[abs(geneList) > 2]
ego <- enrichGO(gene  = gene,
                universe      = names(geneList),
                OrgDb         = org.Hs.eg.db,
                ont           = "BP",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.01,
                qvalueCutoff  = 0.05,
                readable      = TRUE)
```


#Visualization of enrichGO results
```{r}
##### Visualization of enrichGO ######
d <- godata('org.Hs.eg.db', ont="BP")
ego2 <- pairwise_termsim(ego, method="Wang", semData = d)
emapplot(ego2)
emapplot_cluster(ego2)
```
Try GO with all different ont methods parameter 
BP = Biological Processes, CC= Cellular component, MF = Molecular functions

In the following example, we selected fold change above 1 as the differential genes and analyzing their disease association.

```{r}
#### enrich DO #####

library(ggupset)


gene = names(geneList)[abs(geneList) > 1.5]
head(gene)
X = enrichDO(gene,ont = "DO", 
              pvalueCutoff=0.05, 
              pAdjustMethod = "BH", 
              universe = names(geneList),
              minGSSize     = 5,
              maxGSSize     = 500,
              qvalueCutoff  = 0.05,
              readable      = FALSE)

#The readable is a logical parameter, 
#indicates whether the entrezgene IDs will mapping to gene symbols or not
head(X)

#setReadable function helps to convert entrezgene IDs to gene symbols
X <- setReadable(X, 'org.Hs.eg.db')
head(X)

```


#Visualization of enrichDO results
```{r}
## Visualization of enrichDO results ##
barplot(X, showCategory=10)
dotplot(X)

#gene may belong to multiple annotation categories, 
#we developed cnetplot function to extract the complex association between genes and diseases
cnetplot(X, categorySize="pvalue", foldChange=geneList)

#upsetplot is an alternative to 
#cnetplot for visualizing the complex association between genes and diseases.

upsetplot(X)

```

## Enrichment Map ##
Enrichment map organizes enriched terms into a network with edges connecting overlapping gene sets. In this way, mutually overlapping gene sets are tend to cluster together, making it easy to identify functional modules.

```{r}
X2 <- pairwise_termsim(X)
emapplot(X2, showCategory = 10, layout = "star")
```

#Enrichment of NCG
Network of Cancer Gene (NCG)3 is a manually curated repository of cancer genes.NCG release 5.0 (Aug. 2015) collects 1,571 cancer genes from 175 published studies. 
DOSE supports analyzing gene list and determine whether they are enriched in genes known to be mutated in a given cancer type.

```{r}
######## enrichNCG function ####

gene2 <- names(geneList)[abs(geneList) < 3]
ncg <- enrichNCG(gene2)
head(ncg)
```

#Disease Gene Association
The enrichment analysis of disease-gene associations is supported by the enrichDGN function to determine whether the genes have associations with any known diseases 

```{r}
dgn <- enrichDGN(gene)
head(dgn)

```

#Gene set enrichment analysis

```{r}
####### Gene set enrichment analysis GSEA Plot ######
gsecc <- gseGO(geneList=geneList, ont="CC", OrgDb=org.Hs.eg.db, verbose=F)
head(summary(gsecc))
gseaplot(gsecc, geneSetID="GO:0000779")

```

#KEGG Enrichment Analysis

```{r}
library(clusterProfiler)

## KEGG pathway over-representation analysis
data(geneList, package="DOSE")
gene <- names(geneList)[abs(geneList) > 2]

kk <- enrichKEGG(gene         = gene,
                 organism     = 'hsa',
                 pvalueCutoff = 0.05)
head(kk)

## KEGG module over-representation analysis
#KEGG Module is a collection of manually defined function units. In some situation,
#KEGG Modules have a more straightforward interpretation

mkk <- enrichMKEGG(gene = gene,
                   organism = 'hsa',
                   pvalueCutoff = 1,
                   qvalueCutoff = 1)
head(mkk)                   

## KEGG module gene set enrichment analysis ##

mkk2 <- gseMKEGG(geneList = geneList,
                 organism = 'hsa',
                 pvalueCutoff = 1)
head(mkk2)


####### Visualize enriched KEGG pathways #########

## To view the KEGG pathway,use the browseKEGG function,
#which will open a web browser and highlight enriched genes.

browseKEGG(kk, 'hsa04110')

###use the pathview() function from the pathview to visualize enriched KEGG 
##pathways identified by the clusterProfiler package

library("pathview")
hsa04110 <- pathview(gene.data  = geneList,
                     pathway.id = "hsa04110",
                     species    = "hsa",
                     limit      = list(gene=max(abs(geneList)), cpd=1))



```



